Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes

### 2024-12-15 paper.qmd
most changes

This study examines the relationships between socioeconomic, health, and education indicators in U.S. counties using data from the American Community Survey, the USDA Food Environment Atlas, small area income and poverty estimates, and other sources. We will explore regional influences and predictors of life expectancy, and also identify socioeconomic factors associated with school funding adequacy. By analyzing differences in life expectancy and education funding across socioeconomic and regional contexts, this study aims to gain insight into public well-being and economic inequality across the U.S. Through exploratory data analysis and regression modeling with interaction terms, we identify significant regional disparities in both life expectancy and school funding adequacy. The findings highlight that regions, particularly the West and Northeast, tend to have better health and education outcomes, emphasizing the critical role of location and economic conditions in shaping societal well-being.

@@

Life expectancy and school funding are important indicators of public well-being because they capture fundamental aspects of health, education, and social equity that are essential for the development of individuals and communities. Life expectancy is an important health indicator that reflects not only the overall health of a society, but also its stability, economic resilience, and levels of inequality. According to "The Root Causes of Health Inequity - Communities in Action" report from the National Institutes of Health, we can infer that differences in life expectancy between regions and socioeconomic groups often reveal systemic inequalities, with marginalized communities experiencing shorter life spans due to limited access to health care, healthy living environments, and economic opportunities.^1^ Similarly, adequately funded schools ensure that students have equal access to learning opportunities, serving as a crucial investment in social mobility. The Stanford Center for Opportunity Policy in Education highlights that "the unequal allocation and inadequate levels of resources in schools and communities \[are\] at the heart of many gaps in student opportunity."^2^ This emphasizes the importance of equitable and adequate school funding in closing opportunity gaps and empowering students from disadvantaged backgrounds to achieve academic and economic success. Consequently, it can be inferred that underfunded schools, often situated in economically disadvantaged or segregated areas, exacerbate educational disparities and limit students' ability to break the cycle of poverty. These gaps are rooted in systemic inequalities, such as residential segregation, which concentrates poverty, limits access to quality education and healthcare, and perpetuates income and health disparities across generations. By addressing these inequalities, policymakers can foster fairer, more prosperous communities and ensure that everyone benefits from improved health and education systems.

For this purpose, we focus our analysis on two key research questions. First, we investigate how socioeconomic factors—residential segregation, access to healthy foods, and health insurance coverage—impact average life expectancy. **We seek to answer the question, "What is the impact of residential segregation, access to healthy foods, and health insurance coverage on a county's average life expectancy?".** This analysis includes a comparison across different regions to identify variations in the relationship between residential segregation and life expectancy. Second, we evaluate how residential segregation and region influence school funding adequacy. **We seek the answer the question, "Does residential segregation and region impact the likelihood of a school being adequately funded?"**. This question allows us to assess how the racial makeup and geographic location—given that each region of the U.S. is characterized by differing political and cultural norms—affect a county's educational investment. The observational unit for this analysis is the U.S. county.

@@

Variables of interest were selected *a priori* based on their perceived relevance to the two outcomes we wanted to better understand: average life expectancy and school funding adequacy. While nutrition and access to quality healthcare are inherently linked to general wellbeing, it is worthwhile to assess how influential the environmental and financial aspects of these factors are on our tangible outcome. Therefore, we selected two key predictors for our life expectancy model: the percentage of the population with limited access to healthy foods and the percentage of uninsured adults in a county. The former is defined by the US Department of Agriculture as "the percentage of the \[county\] population that is low-income and does not live close to a grocery store", where "close" is defined differently for rural and nonrural areas.^6^ The latter variable represents the percentage of adults in a county younger than 65 who lack health insurance. We also were interested in seeing how residential segregation would impact life expectancy among these predictors, given existing studies on the negative health effects of segregation, and whether this relationship would exhibit any variations across different regions. This is why we chose to include residential segregation index values. The residential segregation index, used by the County Health Rankings & Roadmaps program, measures how evenly two groups, such as Black and white residents, are distributed across geographic areas, with values ranging from 0 (complete integration) to 100 (complete segregation).^7^

*Income is known to have a strong relationship with.*^8^

School funding adequacy is defined by the County Health Rankings program as "the average dollar gap between actual per-pupil spending and the amount needed for students to achieve national average test scores, considering districts' varying equity-based needs."^9^ To explore how racial diversity interface with variations in school funding, we selected residential segregation to be a part of this regression. In addition, given that state policy has a large influence on school administration and the determination of school needs, and that states in the same region tend to have similar political leanings and cultural beliefs, we wanted to analyze how much of a relationship exists between region and actual school funding levels.

Because a major source of public school funding is tax revenue, median property tax and median household income were also included in the school funding model as potential confounders. This allowed us to assess whether the effects of segregation and region on funding existed outside of the effects of known economic factors.

@@

After this, remaining missing values in all variables were imputed using the mice package and predictive mean matching to generate one complete dataset. The original school funding adequacy numerical variable was first imputed, and then converted into binary values for subsequent analysis.

@@

For our second research question, a binary logistic regression model was fit with school funding adequacy (categorical) as the outcome and residential segregation index, region, median property tax, and median household income as predictors. The model was assessed using a confusion matrix, an ROC curve, and a comparison of deviance between the full model and a reduced model that excluded region as a predictor.

@@

Table 1 displays summary statistics for the predictors in the second model, stratified by school funding adequacy. School funding adequacy counts show that around 66% of the counties in our analysis had underfunded schools. Additionally, the number of counties per region indicate an imbalance between categories, with the South region being over represented (holding 60% of all counties), and the East and Northeast holding around 9% of the counties each. Table A1 in the Appendix provides summary statistics for all variables, and shows that both the mean and median for life expectancy are approximately 75 years.

@@

| Variable | **Inadequate School Funding** | **Adequate School Funding** |
|----|----|----|
| Residential Segregation Index: Mean (SD) | 46 (17) | 56 (14) |
| Region-West: Count (%) | 114 (7.3%) | 89 (10.8%) |
| Region-Northeast: Count (%) | 3 (0.2%) | 200 (24.3%) |
| Region-Midwest: Count (%) | 267 (17.1%) | 285 (34.7%) |
| Region-South: Count (%) | 1173 (75.3%) | 248 (30.2%) |
| Median Property Tax, USD: Mean (SD) | 1119 (763) | 2282 (1648) |
| Median Household Income, USD: Mean (SD) | 56513 (13200) | 72218 (18255) |

=======
| Variable                                 | **Inadequate School Funding** | **Adequate School Funding** |
|--------------------------------|---------------------|--------------------|
| Residential Segregation Index: Mean (SD) | 46 (17)                       | 56 (14)                     |
| Region-West: Count (%)                   | 114 (7.3%)                    | 89 (10.8%)                  |
| Region-Northeast: Count (%)              | 3 (0.2%)                      | 200 (24.3%)                 |
| Region-Midwest: Count (%)                | 267 (17.1%)                   | 285 (34.7%)                 |
| Region-South: Count (%)                  | 1173 (75.3%)                  | 248 (30.2%)                 |
| Median Property Tax, USD: Mean (SD)      | 1119 (763)                    | 2282 (1648)                 |
| Median Household Income, USD: Mean (SD)  | 56513 (13200)                 | 72218 (18255)               |

```{r}
# Plot for Q2 EDA
ggpairs(
  completed_data,
  columns = c("Residential.Segregation.Index", "Median.Household.Income", "Median.Prop.Tax"),
  aes(color = factor(School.Funding.Cat,
                     labels = c("Inadequate", "Adequate")))
) +
  labs(
    title = "Pairwise Plots of Predictors Colored by School Funding Adequacy",
    color = "School Funding Adequacy"
  ) +
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    strip.text.x = element_text(size = 5),
    strip.text.y = element_text(size = 3)
  )

```
### Question 1

@@

Based on the multiple linear regression results, residential segregation, percentage of uninsured adults, region, and median household income all emerged with statistically significant effects on life expectancy. However, the coefficients for most of these variables were small. Specifically, holding all other variables constant, for every 1 unit increase in percentage of uninsured adults, a county's average life expectancy increases by 0.032 years, which is surprising and suggests the effect of confounding variables not included in our study like ???. For every \$10,000 increase in median household income, life expectancy increases by 1.25 years on average. The impact of region on life expectancy was notable: life expectancy is on average 1.96 years higher in the West region compared to the South, 2.92 years higher in the Northeast compared to the South, and 1.92 years higher in the Midwest compared to the South, even when controlling for income, uninsured population, and racial diversity. This suggests that potentially the quality of healthcare and public services, which differ substantially from region to region, could be a big driver in life expectancy even outside of income (how to word / explain this better)???

The percentage of population with limited access to healthy foods predictor was not statistically significant, suggesting that this environmental factor may not have an independent effect on life expectancy outside of an economic factor like income. In terms of residential segregation, for every 1 unit increase in a county's index value (corresponding to increased segregation), its average life expectancy increases by 0.0071 years in the South, increases by 0.0068 years in the West, decreases by 0.0055 years in the Northeast, and decreases by 0.0004 years in the Midwest, as seen in Figure 1. However, these differences are not statistically significant given the high p-values of all residential segregation-region interaction terms. We confirmed this result with a nested f-test, comparing a full model to a reduced model that excluded the interaction term between region and residential segregation. This did not show a statistically significant difference in RSS (*p*=0.624), indicating that the interaction between region and residential segregation does not add meaningful explanatory power to our model, and that we did not find enough evidence to support regional variations in the relationship between segregation and life expectancy (???? is this fine???? or am I saying something incorrect???). Instead, based on residential segregation's coefficient estimate in the model, life expectancy does appear to increase overall with an increase in residential segregation, however minor this increase may be.

```{r, echo=FALSE, warning = FALSE, include=FALSE}
# MLR
# Outcome = life expectancy 
# Predictors = residential segregation, limited access to healthy foods, uninsured adults, region_fac as interaction with residential segregation (or any other variable), household income

# To get a general sense of what the data tells us, we use the orignial dataset with missing values.

mlrmodel <- lm(
  Life.Expectancy ~ Residential.Segregation.Index * Region_fac +
  X..Limited.Access.to.Healthy.Foods +
  X..Uninsured.Adults + Median.Household.Income,
  data = combined)

# Renaming the output names 
renamed <- c(
   "Residential Segregation Index (RSI)" = "Residential.Segregation.Index",
  "Percentage With Limited Access to Healthy Foods" = "X..Limited.Access.to.Healthy.Foods",
  "Percentage of Uninsured Adults"= "X..Uninsured.Adults",
  "Region-West" = "Region_facWest", 
  "Region-Northeast"= "Region_facNortheast",
  "Region-Midwest"= "Region_facMidwest",
  "RSI*Region-West" = "Residential.Segregation.Index:Region_facWest",
  "RSI*Region-Northeast" = "Residential.Segregation.Index:Region_facNortheast",
  "RSI*Region-Midwest" = "Residential.Segregation.Index:Region_facMidwest",
  "Median Household Income" = "Median.Household.Income")

modelsummary(mlrmodel, fmt = fmt_significant(2),
             coef_rename = renamed, 
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA,
             )

*Figure 1: Segregation & Life Expectancy Across Different U.S. Regions*

@@

The adjusted R-squared value for the model was 0.6065, indicating that a majority of the variability in life expectancy was explained by the socioeconomic and racial predictors included the model. In conducting diagnostics on the multilinear regression model, we first checked to see if the model met assumptions of linearity, independence, normality, and homoscedasticity. The residuals vs. fitted plot showed relatively equal variance and no discernible pattern, meaning that the linearity and homoscedasticity assumptions were reasonably met. In addition, because each observation in the dataset represented a unique county in the U.S., independence could be reasonably assumed. Assessing the Q-Q Plot for normality, the residual distribution was approximately normal, despite a slight divergence on both tails of the plot. Cook's distance was also assessed using the residuals vs. leverage plot; no points were found to be influential. There was a high leverage point (Mineral County, Nevada), and refitting the model without this point yielded no substantial changes in the coefficient estimates for statistically significant predictors; however, the p-value for residential segregation increased from 0.031 to 0.054. Finally, with a VIF \<2 for all predictors, there was little to no multicollinearity among the predictors.

@@

```{r, echo=FALSE, include=FALSE}
#Cooks Distance 
cooksd <- cooks.distance(mlr.imp)

# Plot Cook's Distance. Not sure if this is required or if the fourth plot in diagnostic plot shows the same thing.
plot(cooksd, main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/(nrow(combined) - length(coef(mlr.imp))), col = "red")  # Threshold line

@@

The table below displays the odds ratio results from the logistic regression model for school funding adequacy, with the South as the baseline category for the categorical region variable. While all estimates are statistically significant at the 0.05 significance level, exponentiated regression coefficients from the logistic model demonstrate that the level of residential segregation, the median household income, and median property tax of the county have very minimal impacts on the odds of a school being adequately funded.

@@

In comparing our full model (with predictor variables for residential segregation, median household income, median property tax, and region) to a model with including all of these variables except for region, we found that there is statistical significance in the deviance between the residuals of the full model and reduced. As such, we can conclude that including region as a predictor refines and creates a better fit for our model.

@@

@@

Evaluating our model, we found that the logistic regression model demonstrated a strong overall performance with an accuracy rate of 80.33%. The model had a high sensitivity of 91%, indicating that it accurately predicted 91% of actual positive cases of adequate school funding when evaluated on our dataset. However, it had a 60% specificity rate, highlighting it correctly identified cases of adequate school funding only 60% of the time. Moreover, looking at the ROC curve below, the AUC is 0.858, indicating a moderately strong discriminate ability.

@@

```{r, echo=FALSE}
#Creating the AUC curves
roc_result <- roc(
  completed_data$School.Funding.Cat,
  predict(imp_mods, type = "response"),
  legacy.axes = TRUE,
  plot = TRUE,
  print.thres = 0.5,
  print.auc = FALSE  # Disable automatic AUC
)

@@

By analyzing these two research questions, we observed the significant impact of region on both life expectancy (a health indicator) and school funding (an education indicator), both of which are key components of societal well-being. These indices tend to be higher in the West and Northeast, which are more developed regions. The findings provide empirical evidence supporting the significant disparities in life expectancy across regions with differing economic conditions. They also provide insights into social policy investment decisions, highlighting the importance of fair resource allocation to people's well-being.

Our analysis faced limitations due to missing data, causing our dataset to be restricted to 33 states. Additionally, because The County Health Rankings program consolidates the latest from a variety of governmental sources, most of our variables were pulled from different years between 2018 and 2022; we assumed that trends remained largely consistent between these years, treating time as insignificant, but this may not actually be the case.

Future studies could incorporate more social variables to enhance the analysis and investigate the causal relationships between region and outcomes. Expanding the dataset to include more states and addressing missing data would also provide a more comprehensive understanding.

@@

*Table A1: Summary Statistics For Numerical Variables*

| Statistic | Residential Segregation Index (RSI) | Percentage With Limited Access to Healthy Foods | Percentage of Uninsured Adults | Median Household Income, USD | Median Property Tax, USD | Average Life Expectancy |
|----|----|----|----|----|----|----|
| Min | 1.0 | 0.0 | 2.8 | 28972 | 0 | 64.9 |
| Median | 50.3 | 6.5 | 12.6 | 58623 | 1173 | 75.1 |
| Mean | 49.3 | 8.2 | 14.1 | 61940 | 1521 | 75.2 |
| Max | 93.8 | 62.6 | 47.1 | 167605 | 9788 | 98.9 |
| Standard Deviation | 16.5 | 6.9 | 6.6 | 16877.8 | 1274.7 | 3.1 |

>>>>

This study examines the relationships between socioeconomic, health, and education indicators in U.S. counties using data from the American Community Survey, the USDA Food Environment Atlas, Small Area Income and Poverty Estimates, and other sources.We analyze how factors such as residential segregation, health insurance coverage, access to healthy foods, and region influence life expectancy, and how residential segregation and region affect school funding adequacy. Our findings reveal that life expectancy is influenced by residential segregation and uninsured rates, though these effects are minor compared to the more substantial influence of household income and regional differences. Life expectancy averages are highest in the Northeast, followed by the Midwest, West, and South. Region also plays the most significant role out of all included variables in determining the likelihood of school funding adequacy, with schools in the Northeast and Midwest being far more likely to receive adequate funding than those in the South. Other factors, such as residential segregation, property taxes, and household income, show statistically significant but smaller effects. These results highlight the complex interplay between socioeconomic variables and geographic location in shaping public well-being. By identifying key disparities in health and education, this study underscores the importance of equitable resource allocation and targeted policies to address regional and socioeconomic inequalities.

@@

Life expectancy and school funding are important indicators of public well-being because they capture fundamental aspects of health, education, and social equity that are essential for the development of individuals and communities. Life expectancy is an important health indicator that reflects not only the overall health of a society, but also its stability, economic resilience, and levels of inequality. While access to quality healthcare and nutritious foods are drivers of well-being, where you live and the nature of the community you live in likely also have an influence on health outcomes such as life expectancy. According to "The Root Causes of Health Inequity - Communities in Action" report from the National Institutes of Health, we can infer that differences in life expectancy between regions and socioeconomic groups often reveal systemic inequalities, with marginalized communities experiencing shorter life spans due to limited access to health care, healthy living environments, and economic opportunities.^1^ When it comes to educational indicators, adequately funded schools ensure that students have equal access to learning opportunities, serving as a crucial investment in social mobility. The Stanford Center for Opportunity Policy in Education highlights that "the unequal allocation and inadequate levels of resources in schools and communities \[are\] at the heart of many gaps in student opportunity."^2^ This emphasizes the importance of equitable and adequate school funding in closing opportunity gaps and empowering students from disadvantaged backgrounds to achieve academic and economic success. By better understanding the correlates of school funding adequacy, we can take a major step towards implementing policy that targets the areas and communities most affected by this issue.

For this purpose, we focus our analysis on two key research questions. First, we investigate how socioeconomic factors impact average life expectancy. **We seek to answer the question, "What is the impact of residential segregation, access to healthy foods, and health insurance coverage on a county's average life expectancy?".** This analysis includes a comparison across different regions to identify variations in the relationship between residential segregation and life expectancy. Second, **we seek to answer the question, "Do residential segregation and region impact the likelihood of a school being adequately funded?"**. This question allows us to assess how the racial makeup and geographic location—given that each region of the U.S. is characterized by differing political and cultural norms—affect a county's educational investment. The observational unit for this analysis is the U.S. county.

@@

Variables of interest were selected a priori based on their perceived relevance to the two outcomes we wanted to better understand: average life expectancy and school funding adequacy. While nutrition and access to quality healthcare are inherently linked to general wellbeing, it is worthwhile to assess how influential the environmental and financial aspects of these factors are on our tangible outcome. Therefore, we selected two key predictors for our life expectancy model: the percentage of the population with limited access to healthy foods and the percentage of uninsured adults in a county. The former is defined by the US Department of Agriculture as "the percentage of the \[county\] population that is low-income and does not live close to a grocery store", where "close" is defined differently for rural and nonrural areas.^6^ The latter variable represents the percentage of adults in a county younger than 65 who lack health insurance. We also were interested in seeing how residential segregation would impact life expectancy among these predictors, given existing studies on the negative health effects of segregation, and whether this relationship would exhibit any variations across different regions. This is why we chose to include residential segregation index values as a predictor. The residential segregation index, used by the County Health Rankings & Roadmaps program, measures how evenly two groups, such as Black and white residents, are distributed across geographic areas, with values ranging from 0 (complete integration) to 100 (complete segregation).^7^

Past studies have established a correlation between income and life expectancy.^8^ Because we wanted to control for its effects and examine how insurance, healthy food access, segregation, and region influence life expectancy even within a given income bracket, we selected income as our final predictor for the first model.

School funding adequacy is defined by the County Health Rankings program as "the average dollar gap between actual per-pupil spending and the amount needed for students to achieve national average test scores, considering districts' varying equity-based needs."^9^ To explore how racial diversity interface with variations in school funding, we selected residential segregation to be a part of this regression. In addition, given that state policy has a large influence on school administration and the determination of school needs, and that states in the same region tend to have similar political leanings and cultural beliefs, we wanted to analyze how much of a relationship exists between region and actual school funding levels. Because a major source of public school funding is tax revenue, median property tax and median household income were also included in the school funding model as potential confounders. This allowed us to assess whether the effects of segregation and region on funding existed outside of the effects of known economic factors.

@@

After this, the remaining missing values in all variables were imputed using the mice package and predictive mean matching to generate one complete dataset. The original school funding adequacy numerical variable was first imputed, and then converted into binary values for subsequent analysis.

@@

For our second research question, a binary logistic regression model was fit with school funding adequacy (categorical) as the outcome and residential segregation index, region, median property tax, and median household income as predictors. The model was assessed using a confusion matrix, an ROC curve, and a comparison of deviance between the full model and a reduced model that excluded region as a predictor. VIF and Cook's distance were again used to evaluate multicollinearity and influential points.

@@

Table 1 displays summary statistics for the predictors in the second model, stratified by school funding adequacy. Table A1 in the Appendix provides summary statistics for all variables, and shows that both the mean and median for life expectancy are approximately 75 years. School funding adequacy counts show that around 66% of the counties in our analysis had underfunded schools. Additionally, the number of counties per region indicate an imbalance between categories, with the South region being over represented (holding 60% of all counties), and the East and Northeast holding around 9% of the counties each.

@@

| Variable                                 | **Inadequate School Funding** | **Adequate School Funding** |
|------------------------------|----------------------|--------------------|
| Residential Segregation Index: Mean (SD) | 46 (17)                       | 56 (14)                     |
| Region-West: Count (%)                   | 114 (7.3%)                    | 89 (10.8%)                  |
| Region-Northeast: Count (%)              | 3 (0.2%)                      | 200 (24.3%)                 |
| Region-Midwest: Count (%)                | 267 (17.1%)                   | 285 (34.7%)                 |
| Region-South: Count (%)                  | 1173 (75.3%)                  | 248 (30.2%)                 |
| Median Property Tax, USD: Mean (SD)      | 1119 (763)                    | 2282 (1648)                 |
| Median Household Income, USD: Mean (SD)  | 56513 (13200)                 | 72218 (18255)               |

```{r, echo=FALSE, include=FALSE}
# Plot for Q2 EDA
ggpairs(
  completed_data,
  columns = c("Residential.Segregation.Index", "Median.Household.Income", "Median.Prop.Tax"),
  aes(color = factor(School.Funding.Cat,
                     labels = c("Inadequate", "Adequate")))
) +
  labs(
    title = "Pairwise Plots of Predictors Colored by School Funding Adequacy",
    color = "School Funding Adequacy"
  ) +
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    strip.text.x = element_text(size = 5),
    strip.text.y = element_text(size = 3)
  )

### Question 1

@@

Based on the multiple linear regression results, residential segregation, percentage of uninsured adults, region, and median household income all showed statistically significant effects on life expectancy. Specifically, for every \$10,000 increase in median household income, life expectancy increases by 1.25 years on average. In addition, the impact of region on life expectancy was substantial: life expectancy is on average 1.96 years higher in the West region compared to the South, 2.92 years higher in the Northeast compared to the South, and 1.92 years higher in the Midwest compared to the South, even when controlling for income, uninsured population, and racial diversity. This suggests, income as a factor withstanding, the differing quality of and access to healthcare and public services found in each region can have a significant impact on life expectancy.

Surprisingly, holding all other variables constant, for every 1 unit increase in percentage of uninsured adults in a county, a county's average life expectancy increases by 0.032 years. While statistically significant, many studies covering the relationship between an individual’s health and having insurance coverage have proven its positive correlation, suggesting our results may benefit from a closer look. In assessing our dataset, as seen in Appendix 1A, we find that our dataset contains more observations for counties with low uninsured individuals, or equivalently, counties where most people are covered. As the timeframe of our dataset is 2018-2022, it is likely that the aforementioned observations require greater scrutiny of the significance of the impact of this variable on average life expectancy of a county.

The percentage of population with limited access to healthy foods predictor was not statistically significant in our model, suggesting that the effect of this environmental factor is minor or intertwined with economic variables, such as income, rather than being a direct or standalone influence. In terms of residential segregation, for every 1 unit increase in a county's index value (corresponding to increased segregation), its average life expectancy increases by 0.0071 years in the South, increases by 0.0068 years in the West, decreases by 0.0055 years in the Northeast, and decreases by 0.0004 years in the Midwest, as seen in Figure 1. However, these regional differences are not statistically significant at the 0.05 significance level given the high p-values of all residential segregation-region interaction terms. We confirmed this result with a nested F-test, comparing a full model to a reduced model that excluded the interaction term between region and residential segregation. This did not show a statistically significant difference in RSS (p=0.624), indicating that the interaction between region and residential segregation does not add meaningful explanatory power to our model, Consequently, we did not find sufficient evidence to support regional variations in the relationship between segregation and life expectancy. Instead, based on residential segregation's coefficient estimate in the model, life expectancy does appear to increase overall with an increase in residential segregation, however minor this increase may be. One possible theory for this observation is that counties with more racial uniformity might experience less racial conflict and more social cohesion, especially for minority communities.

*Figure 1: Segregation vs. Life Expectancy Across Different U.S. Regions*

@@

The adjusted R-squared value for the model was 0.6065, indicating that a majority of the variability in life expectancy was explained by the socioeconomic and racial predictors included in the model. In conducting diagnostics on the multilinear regression model, we first checked to see if the model met assumptions of linearity, independence, normality, and homoscedasticity. The residuals vs. fitted plot showed relatively equal variance and no discernible pattern, meaning that the linearity and homoscedasticity assumptions were reasonably met. In addition, because each observation in the dataset represented a unique county in the U.S., independence could be reasonably assumed. Assessing the Q-Q Plot for normality, the residual distribution was approximately normal, despite a slight divergence on both tails of the plot. Cook's distance was also assessed using the residuals vs. leverage plot; no points were found to be influential. There was a high leverage point (Mineral County, Nevada), and refitting the model without this point yielded no substantial changes in the coefficient estimates for statistically significant predictors; however, the p-value for residential segregation increased from 0.031 to 0.054. Finally, with a VIF \<2 for all predictors, there was little to no multicollinearity among the predictors.

@@

@@

Table 3 below displays the odds ratio results from the logistic regression model for school funding adequacy, with the South as the baseline category for the categorical region variable. While all estimates are statistically significant at the 0.05 significance level, exponentiated regression coefficient estimates from the logistic model demonstrate that the level of residential segregation, the median household income, and median property tax of the county have very minimal impacts on the odds of a school being adequately funded.

@@

In comparing our full model (with predictor variables for residential segregation, median household income, median property tax, and region) to a model including all of these variables except for region, we found that there is statistical significance in the deviance between the residuals of the full model and reduced. As such, we can conclude that including region as a predictor refines and creates a better fit for our model.

@@

*Figure 2: Relationship Between Residential Segregation, Region, & School Funding Adequacy*

```{r}
ggplot(completed_data, aes(x=Region_fac, y = Residential.Segregation.Index, color=factor(School.Funding.Cat))) +
  geom_boxplot() +
  labs(x = "U.S. Region", y='Residential Segregation Index', color='School Funding Adequacy', title='Residential Segregation Stratified by Region & School Funding') +
  scale_color_hue(labels = c("Inadequate", "Adequate"))+ 
  theme(plot.margin = margin(t = 20, r = 30, b = 20, l = 10),
axis.title.x = element_text(size = 8),
axis.title.y = element_text(size = 8),
title=element_text(size=9))
```

@@

To assess logistic regression assumptions, we first plotted each numerical predictor against the log odds of school funding adequacy. All plots showed a mostly linear relationship, indicating that the linearity assumption was largely met; interestingly, different linear trends were observed within each plot based on region, pointing to potential interaction effects between region and the three numerical predictors. GVIF for all predictors ranged between 1 and 3, suggesting that there could be moderate multicollinearity between predictors, but not enough to warrant a reconsideration of included variables. Cook's distance was also less than 0.05 for all points, meaning that there were no highly influential observations in the model. Evaluating our model, we found that the logistic regression model demonstrated a strong overall performance with an accuracy rate of 80.33%. The model had a high sensitivity of 91%, indicating that it accurately predicted 91% of actual positive cases of adequate school funding when evaluated on our dataset. However, it had a 60% specificity rate, highlighting it correctly identified cases of inadequate school funding only 60% of the time. Overall, looking at the ROC curve below, the AUC is 0.858, indicating a moderately strong discriminate ability.

```{r, include=FALSE, echo=FALSE}
plot(completed_data$Residential.Segregation.Index, predict(imp_mods), col = completed_data$Region_fac)
plot(completed_data$Median.Household.Income, predict(imp_mods), col = completed_data$Region_fac)
plot(completed_data$Median.Prop.Tax, predict(imp_mods), col = completed_data$Region_fac)
```

```{r, include=FALSE, echo=FALSE}
#Cooks Distance 
cooksd <- cooks.distance(imp_mods)

# Plot Cook's Distance
plot(cooksd, main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/(nrow(completed_data) - length(coef(imp_mods))), col = "red")  # Threshold line
4/(nrow(data) - length(coef(imp_mods)))
```

```{r, echo=FALSE, include=FALSE}
vif(imp_mods)
plot(imp_mods)
```

@@

```{r, echo=FALSE, include=FALSE}
#Creating the AUC curves
roc_result <- roc(
  completed_data$School.Funding.Cat,
  predict(imp_mods, type = "response"),
  legacy.axes = TRUE,
  plot = TRUE,
  print.thres = 0.5,
  print.auc = FALSE  # Disable automatic AUC
)

@@

In our analysis, we observed the significant impact of the regional location of a county on both its average life expectancy (a health indicator) and school funding adequacy (an educational indicator)--two key components of societal and an individual’s well-being. Life expectancy is highest in the Northeast, followed by the Midwest, West, and South, while school funding adequacy is significantly greater in the Northeast and Midwest compared to the South. Our findings highlight substantial differences in both outcomes between regions, which are often caused by different economic capabilities and different allocations of resources among countries. This study provides empirical evidence that underscores the importance for equitable social policies and targeted investments to address regional disparities and promote well-being across all communities.

Our analysis faced limitations due to missing data, which restricted the dataset to 33 states and reduced the generalizability of our findings. Additionally, some variables were drawn from different years between 2018 and 2022, requiring us to assume that trends remained consistent over a 1–5 year period and that time was an insignificant factor; however, this may not actually be the case.

Future studies could incorporate additional social variables to enhance the analysis and explore causal relationships between region and outcomes. For example, integrating a dataset on employment rates and school performance could offer deeper insights into how socioeconomic factors shape life expectancy and educational funding, complementing existing variables in our models. Another direction for future work is to delve deeper into understanding the surprising findings regarding the effects of uninsured adult percentage and residential segregation on life expectancy. Expanding the dataset to include a temporal scope, and finding more complete sources of data on variables of interest so that no states have to be excluded from the analysis, would also contribute to a more comprehensive understanding.

@@

*Table A1: Summary Statistics For All Variables*

![](images/Screenshot 2024-12-15 at 2.38.35 AM.png)

```{r}
renamed_data <- completed_data %>%
  rename(
    "Residential Segregation Index" = Residential.Segregation.Index,
    "Median Household Income" = Median.Household.Income,
    "Median Property Tax" = Median.Prop.Tax,
    "School Funding Adequacy" = School.Funding.Adequacy, 
    "Average Life Expectancy" = Life.Expectancy,
    "% Limited Access to Healthy Foods" = X..Limited.Access.to.Healthy.Foods,
    "% Uninsured Adults" = X..Uninsured.Adults,
    "Region" = Region_fac,
    "School Funding Binary" =School.Funding.Cat)

datasummary_skim(
  renamed_data,
  title = "Customized Summary Table")
